%% ============================================================
%% introduction.tex  —  Chapter 1
%% ============================================================
\chapter{Introduction}
\label{ch:introduction}

\section{The Search for Habitable Environments Beyond Earth}
\label{sec:search_context}

The question of whether life exists elsewhere in the universe has undergone a
profound transformation over the past three decades, shifting from the domain
of philosophical speculation to empirical scientific investigation.  This
change has been driven by two largely independent but mutually reinforcing
developments: the systematic discovery that life on Earth thrives in conditions
previously considered inimical to biology, and the growing recognition through
robotic spacecraft exploration that chemically and thermodynamically suitable
environments for life are far more widely distributed throughout the solar
system than twentieth-century intuitions suggested.

The progressive characterisation of terrestrial extremophile communities has
fundamentally expanded the envelope of conditions considered compatible with
life.  Microorganisms have been cultured from deep-sea hydrothermal vents at
temperatures exceeding 121\textdegree{}C, from hypersaline brines with water
activities as low as $a_w = 0.60$, from the anoxic pore spaces of deep
crustal rocks where electrons liberated by water--rock reactions provide
metabolic energy in the total absence of sunlight, and from sub-glacial
Antarctic lakes that have been isolated from the atmosphere for millions of
years \citep{Rothschild2001, Cavicchioli2002}.  Each such discovery has
moved the habitable niche boundaries outwards, making it progressively more
difficult to argue that any particular planetary environment is categorically
uninhabitable without specific evidence that the relevant physical and
chemical conditions lie entirely outside the known envelope of life.

In parallel, the outer solar system has revealed a far richer array of
liquid environments than the classical habitable zone framework would predict.
The Europa detection from the Galileo mission \citep{Pappalardo1999} and the
subsequent discovery of active cryovolcanism and ocean chemistry at Enceladus
\citep{Waite2017} established that tidal heating can maintain liquid water
oceans beneath ice shells at heliocentric distances far beyond any reasonable
definition of the water-based habitable zone.  These discoveries shifted the
conceptual framework of solar system habitability from a two-dimensional
picture (terrestrial planets in a narrow orbital band) to a three-dimensional
one encompassing subsurface environments throughout the outer solar system.

Within this broader astrobiological context, Saturn's moon \titan{} occupies
a distinctive and arguably singular position.  It is the only moon in the
solar system possessing a substantial atmosphere --- one that is denser at the
surface than Earth's and dominated by molecular nitrogen, with a structurally
complex photochemistry driven by the ultraviolet flux of the distant Sun.
It is the only known extraterrestrial body to host stable surface liquids
\citep{Stofan2007}, completing a full precipitation-evaporation-lake cycle
that is the structural analogue of the terrestrial hydrological cycle.
It may have accumulated the most complex organic chemical inventory in the
solar system outside Earth itself \citep{Lorenz2008}.  And at depth, it
contains a water--ammonia ocean whose thermodynamic conditions are
increasingly well-constrained by tidal measurements and geophysical
modelling.  The combination of these properties makes \titan{} not
merely a potentially habitable world but a natural laboratory for
prebiotic chemistry operating at planetary scale and across a geological
timescale of $\sim$4.5\,Gyr.

\section{The Physical Environment of Titan}
\label{sec:titan_environment}

\titan{}'s dense nitrogen--methane atmosphere ($\sim$94\% \ce{N2}, $\sim$5.6\% \ce{CH4}; surface pressure $\sim$1.5\,bar, surface temperature $\sim$94\,K; \citealt{Fulchignoni2005,Niemann2005}) drives a methane cycle analogous to Earth's water cycle, with rain, rivers, and polar hydrocarbon lakes.  Ultraviolet photolysis of the \ce{N2}/\ce{CH4} mixture continuously produces tholins at $\sim5\times10^{-14}$\,g\,cm$^{-2}$\,s$^{-1}$ \citep{Lavvas2011}, building a global organic sediment layer estimated at 1--10\,km depth over solar system history \citep{Lorenz2008}.

The surface is geologically diverse: equatorial organic dune fields ($\sim$17\% of the SAR-imaged area), plains ($\sim$65\%), and a north polar sea system --- Kraken Mare, Ligeia Mare, and Punga Mare --- containing an estimated 70{,}000\,km$^3$ of liquid hydrocarbons \citep{Mastrogiuseppe2019,Malaska2025}.  The Lopes et al.\ \citeyearpar{Lopes2019} global geomorphological map, derived from the full Cassini SAR dataset, provides the terrain classification underpinning the organic abundance feature used in this thesis.

The interior harbours a global water--ammonia subsurface ocean at $\sim$55--80\,km depth, inferred from the tidal Love number $k_2 = 0.589\pm0.150$ \citep{Iess2012}.  This ocean provides a potential habitat in its own right and, at past epochs, a source of liquid water during impact-melt events where ocean-derived chemistry contacts the surface organic inventory \citep{Tobie2005}.

\section{Chemical Habitability Scenarios}
\label{sec:prior_theory}

Four habitability pathways have been proposed for \titan{}'s surface.

\begin{enumerate}[itemsep=0em]
    \item The \textit{acetylenotrophy hypothesis} \citep{McKaySmith2005} identifies $\ce{C2H2 + 3H2 -> 2CH4}$ ($\Delta G = -334$\,kJ\,mol$^{-1}$) as a viable energy source at the present epoch, with acetylene and \ce{H2} supplied by atmospheric photolysis; the energetics comfortably exceed the metabolic threshold of terrestrial methanogens even under updated thermochemical estimates \citep{Yanez2024}.

    \item The \textit{azotosome hypothesis} \citep{Stevenson2015} proposes that acrylonitrile (detected by ALMA; \citealt{Palmer2017}) spontaneously forms stable membrane-like vesicles in liquid methane, providing a plausible route to cellular compartmentalisation without aqueous chemistry.


    \item \textit{Post-impact melt chemistry} offers a water-based pathway: large impacts produce water--ammonia--organic melt ponds persisting for $10^3$--$10^4$\,yr during which Strecker synthesis yields amino acids from surface aldehydes, ammonia, and HCN \citep{Neish2010, Madan2026}; this is the primary astrobiological rationale for the \dragonfly{} landing site at Selk crater.


    \item Finally, the \textit{far-future habitability window} \citep{Lorenz1997} arises when red-giant solar luminosity raises \titan{}'s surface above the water--ammonia eutectic (176\,K) for an estimated 200--500\,Myr, creating a global ocean over 16\,Gyr of accumulated organics.

\end{enumerate}

This thesis provides a quantitative characterisation of all four scenarios within a unified Bayesian framework across geological time.

\section{The Cassini--Huygens Mission and Dragonfly}
\label{sec:missions}

The \cassini{}--\huygens{} mission (2004--2017) provided the observational foundation of this thesis through 127 targeted \titan{} flybys, the \huygens{} atmospheric probe descent and landing, and a systematic multi-instrument survey of \titan{}'s surface, atmosphere, and interior.  SAR mapping covered $\sim$67\% of the surface; VIMS near-infrared spectroscopy $\sim$50\%; CIRS thermal mapping the full globe at $\sim$10\textdegree{} resolution.  The mission's key unresolved gap is the absence of a quantitative, spatially-resolved, multi-epoch habitability analysis integrating all data modalities --- which this thesis provides.

NASA's \dragonfly{} rotorcraft \citep{Lorenz2018Dragonfly}, due to arrive in 2034, will traverse $\sim$175\,km from the Belet dune field to Selk crater over a 2.7-year mission.  Chapter~\ref{ch:results} provides a quantitative evaluation of that site selection relative to globally optimal alternatives.

\section{Scope and Research Questions}
\label{sec:scope}

Several preceding studies have attempted to place planetary habitability on
a quantitative footing.  The \textit{Earth Similarity Index}
\citep{Schulze-Makuch2011} scores planetary bodies against Earth on
temperature, density, escape velocity, and surface chemistry, but
produces a single scalar per body and cannot represent spatial or temporal
variation.  \citet{Hendrix2019} reviewed the habitability potential of
icy ocean worlds using qualitative multi-criteria matrices, providing rich
scientific context but no spatial resolution or probabilistic framework.
Logistic regression and machine-learning habitability classifiers
(e.g.\ \citealt{Spiegel2012}) have been applied to exoplanet populations
but require labelled training data that does not exist for Titan.
The Bayesian approach adopted here differs from all of these: it encodes
scientific prior knowledge as probability distributions, updates them
with observational data via conjugate updating, and returns credible
intervals rather than point estimates --- making uncertainty explicit
rather than implicit.  The choice of a Beta-conjugate (Beta-Binomial)
model is motivated by the bounded $[0,1]$ domain of habitability
probability and by the analytical tractability that permits simultaneous
evaluation at all 6.49 million grid pixels without numerical integration
or MCMC.  A Gaussian process or multivariate logistic regression could
in principle capture feature correlations more richly, but would require
either strong covariance priors or prohibitively dense labelled sampling
to estimate the correlation structure from Cassini data alone.  The
Beta-conjugate approach is preferred as the most tractable framework
that preserves physical interpretability at every pixel.

Previous habitability assessments of \titan{} have shared four important
limitations.  

\begin{enumerate}[itemsep=0em]
    \item They are predominantly qualitative or point-specific, lacking
global spatial resolution.  
    \item They treat the present epoch as the only relevant
state, ignoring both the LHB prebiotic-chemistry window and the far-future
red-giant window.  
    \item They do not integrate the full suite of Cassini data
modalities into a unified probabilistic framework.  
    \item They lack the
reproducibility afforded by an openly archived analysis pipeline.

\end{enumerate}

The closest methodological precedent is \citet{Affholder2021}, who applied
Bayesian statistical inference to Cassini INMS plume data from Enceladus,
comparing the observed methane and hydrogen escape rates against models of
abiotic serpentinisation and biotic methanogenesis.  Their finding --- that
the observed methane escape rate is (with a minimal number of assumptions) explained by biological methanogenesis if the prior probability of life is sufficient ---
represents a comparable Bayesian biosignature assessment for a solar system body.  This thesis is directly inspired
by their approach, extended from a single-point plume observation to a full-globe, multi-feature, multi-epoch spatial mapping.

The specific research questions addressed are:
\begin{description}
  \item[RQ1] \textit{What is the spatial distribution of surface habitability
    on present-day \titan{}, and which terrain types, geological features,
    and atmospheric zones produce the highest scores?  Are the data-derived
    feature importances consistent with the physically motivated prior weights?}

  \item[RQ2] \textit{How did habitability differ during the Late Heavy
    Bombardment, when impact-melt ponds provided transient liquid water?
    Which impact structures were the most astrobiologically significant,
    and what are the implications for prebiotic chemistry?}

  \item[RQ3] \textit{How does habitability evolve through the temporal sequence
    from the LHB to the red-giant future?  Is \titan{} potentially the most
    habitable body in the solar system during the red-giant ocean window?}

  \item[RQ4] \textit{Does the derived habitability map independently validate
    the \dragonfly{} landing site at Selk crater, and which instruments
    are best positioned to test the model's predictions?}
\end{description}

Chapter~\ref{ch:methods} presents the data sources, feature extraction,
Bayesian inference model, and temporal methodology.
Chapter~\ref{ch:results} presents and interprets the habitability maps at
each epoch, the feature importances, the temporal evolution, and the
detailed Selk crater analysis.
Chapter~\ref{ch:discussion} discusses the \dragonfly{} mission implications,
biosignature framework, limitations, and future work.

