%% ============================================================
%% main.tex  —  Titan Habitability MPhil Thesis
%% Compile: pdflatex main && bibtex main && pdflatex main && pdflatex main
%% ============================================================
\documentclass[12pt,a4paper,twoside]{report}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
% lmodern fonts not in this TeX installation; using CM
\usepackage[margin=2.5cm,inner=3.2cm]{geometry}
\usepackage{setspace}
% microtype disabled (requires scalable fonts)
% \usepackage{microtype}
\usepackage{emptypage}
\usepackage{amsmath,amssymb,mathtools}
\usepackage[version=4]{mhchem}
\usepackage{textcomp}   % provides \textdegree; do NOT redefine \degree
\usepackage{bm}
\usepackage{graphicx}
\usepackage{float}
\usepackage[labelfont=bf,font=small]{caption}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}
\usepackage{tabularx}
\usepackage[dvipsnames]{xcolor}
\usepackage[
  colorlinks=true,
  linkcolor=MidnightBlue,
  citecolor=ForestGreen,
  urlcolor=Maroon,
  pdfauthor={Chris Meadows},
  pdftitle={Bayesian Surface Habitability Mapping of Titan},
]{hyperref}
\usepackage{cleveref}
\usepackage[round,authoryear,sort&compress]{natbib}
\usepackage{fancyhdr}
\usepackage{enumitem}
\usepackage{siunitx}
\sisetup{separate-uncertainty=true,per-mode=symbol}

% Figure paths — all pipeline outputs are relative to ../titan_pipeline/
\graphicspath{{../titan_pipeline/}}

% Reduce the excessive vertical space before chapter titles
\usepackage{titlesec}
\titleformat{\chapter}[hang]
  {\normalfont\LARGE\bfseries}{\thechapter}{1em}{}
\titlespacing*{\chapter}{0pt}{20pt}{16pt}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE]{\small\itshape\leftmark}
\fancyhead[RO]{\small\itshape\rightmark}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% Custom commands
\newcommand{\titan}{Titan}
\newcommand{\cassini}{\textit{Cassini}}
\newcommand{\huygens}{\textit{Huygens}}
\newcommand{\dragonfly}{\textit{Dragonfly}}
\newcommand{\featurevec}{\mathbf{f}}
\newcommand{\gya}{\,\text{Gya}}
\newcommand{\myr}{\,\text{Myr}}
\DeclareMathOperator{\Beta}{Beta}

\begin{document}
\doublespacing

%% ----- Title page -------------------------------------------------------
\begin{titlepage}
\centering
\vspace*{1.8cm}
{\LARGE\bfseries Bayesian Surface Habitability Mapping of Titan}\\[0.7em]
{\large\itshape from the Late Heavy Bombardment to the Solar Red-Giant Phase}
\vspace{2.5cm}

{\normalsize A report submitted for the degree of}\\[0.4em]
{\large\bfseries Master of Philosophy}\\[0.2em]
{\large Planetary Science and Life in the Universe}
\vspace{2.0cm}

{\large\itshape Chris Meadows}\\[0.4em]
{\normalsize \texttt{cm10004@cam.ac.uk}}\\[0.2em]
{\normalsize Institute of Astronomy, University of Cambridge}
\vspace{2.0cm}

{\large March 2026}
\vfill
\rule{0.5\textwidth}{0.4pt}\\[0.4em]
{\footnotesize All code, processed data products, and static habitability
  maps are openly archived at
  \url{https://github.com/chrimblechrumble/pslu-p3}.}
\end{titlepage}

%% ----- Abstract ---------------------------------------------------------
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Saturn's moon \titan{} is unique amongst solar system bodies in simultaneously
possessing a dense nitrogen-dominated atmosphere, a global methane--ethane
precipitation and lake cycle that is structurally analogous to the terrestrial
hydrological cycle, the largest inventory of complex organic chemistry in the
solar system outside Earth, confirmed polar hydrocarbon seas containing an
estimated $\sim$70{,}000\,km$^3$ of liquid methane--ethane--nitrogen, and a
deep interior water--ammonia ocean inferred from tidal deformation
measurements.  This convergence of chemical feedstock, cycling solvent, and
liquid environments makes \titan{} not merely a potentially habitable world
but a natural laboratory for prebiotic chemistry at planetary scale.

This thesis presents a quantitative, spatially-resolved probabilistic
framework for assessing \titan{}'s surface and near-surface habitability at
five temporal configurations spanning the Late Heavy Bombardment
($-3.5\gya$), a lake-formation transition epoch ($-1.0\gya$), the
present-day \cassini{} epoch, a near-future solar warming window ($+0.25\gya$),
and the red-giant future ($+5.9\gya$).  Eight observational
habitability-proxy features are derived from the complete \cassini{} dataset
--- synthetic aperture radar (SAR), near-infrared spectroscopy (VIMS),
thermal mapping (CIRS), radar altimetry, and the Lopes et al.\ (2019) global
geomorphological map --- and combined through a Bayesian Beta-conjugate
posterior update to yield probability maps $P(\text{habitable}\mid\featurevec)$
at \SI{4490}{\metre\per\text{px}} resolution across the full globe.

At the present epoch, habitability is highest along the margins of the north
polar hydrocarbon seas, where posterior means reach $P(H)\approx0.55$--0.63.
During the Late Heavy Bombardment, impact-melt ponds at major crater sites
constituted the dominant habitability zones.  In the far future, solar
red-giant expansion will produce a near-globally habitable water--ammonia
ocean over an extraordinary accumulated organic substrate.

The analysis independently validates the NASA \dragonfly{} mission's selection
of Selk crater, which ranks in the 98th percentile of present-epoch and 99th
percentile of past-epoch habitability.  The instrument suite is shown to be
well-matched to the dominant habitability drivers at the target site.

\vspace{1em}
\noindent\textbf{Keywords:} \titan{}; habitability; Bayesian inference;
\cassini{}; astrobiology; prebiotic chemistry; multi-epoch; temporal
evolution; \dragonfly{}

%% ----- Acknowledgements -------------------------------------------------
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

The author gratefully acknowledges the NASA/ESA \cassini{}--\huygens{}
mission team, whose thirteen years of sustained observation produced the
extraordinary dataset underpinning every quantitative result presented here.
The Lopes et al.\ (2019) global geomorphological shapefiles were made
available by Dr Rosaly Lopes and colleagues at the Jet Propulsion Laboratory.
The Corlies et al.\ (2017) globally interpolated topographic model and the
Miller et al.\ (2021) global fluvial channel map were obtained through the
Hayes Research Group at Cornell University.  The Seignovert et al.\ (2019)
VIMS+ISS global mosaic is archived at CaltechDATA.

\cleardoublepage

%% ----- Chapters ---------------------------------------------------------
\input{introduction}
\input{methods}
\input{results}
\input{discussion}

%% ----- Bibliography -----------------------------------------------------
\addcontentsline{toc}{chapter}{References}
\bibliographystyle{apalike}
\bibliography{references}

%% ----- Appendices -------------------------------------------------------
\appendix

\chapter{Physical Parameters for Titan}
\label{app:constants}

\begin{table}[H]
\centering
\caption{Physical constants and observationally-constrained parameters for
\titan{} used throughout this thesis.}
\label{tab:constants}
\begin{tabular}{p{5.2cm}p{3.5cm}p{4.0cm}}
\toprule
\textbf{Parameter} & \textbf{Value} & \textbf{Reference} \\
\midrule
Mean radius              & 2575.0\,km              & IAU 2015 \\
Surface gravity          & 1.352\,m\,s$^{-2}$      & \citet{Zebker2009} \\
Surface pressure         & $1.467\pm0.001$\,bar     & \citet{Fulchignoni2005} \\
Surface temperature      & $93.65\pm0.25$\,K        & \citet{Fulchignoni2005} \\
\ce{N2} fraction (surface) & $\approx94.2\%$       & \citet{Niemann2005} \\
\ce{CH4} fraction (surface) & $5.65\pm0.18\%$      & \citet{Niemann2005} \\
Tidal Love number $k_2$  & $0.589\pm0.150$          & \citet{Iess2012} \\
Saturn orbital period    & 29.46\,yr               & --- \\
Axial obliquity          & 26.73\textdegree         & --- \\
Solar flux at \titan{}   & $\approx14.9$\,W\,m$^{-2}$ & at 9.54\,AU \\
Organic deposition rate  & $\sim5\times10^{-14}$\,g\,cm$^{-2}$\,s$^{-1}$ & \citet{Lavvas2011} \\
North polar sea volume   & $\sim70{,}000$\,km$^3$  & \citet{Mastrogiuseppe2019} \\
Water--ammonia eutectic  & 176\,K at 32.6\,wt\% \ce{NH3} & \citet{Grasset2005} \\
Subsurface ocean depth   & $\sim55$--80\,km         & \citet{Tobie2005} \\
\bottomrule
\end{tabular}
\end{table}

\chapter{Bayesian Prior Specification}
\label{app:priors}

\begin{table}[H]
\centering
\caption{Prior means $\mu_i$ and weights $w_i$ for all eight habitability
features at the present (Cassini) epoch.  Global prior mean
$\mu_0 = \sum_i w_i\mu_i = 0.331$.  Concentration parameter $\kappa = 5$
is applied uniformly.}
\label{tab:priors}
\begin{tabular}{llrrl}
\toprule
\textbf{Feature} & \textbf{Symbol} & \textbf{$w_i$} & \textbf{$\mu_i$} & \textbf{Key ref.} \\
\midrule
Liquid hydrocarbon         & $f_1$ & 0.25 & 0.020 & \citet{Hayes2008} \\
Organic abundance          & $f_2$ & 0.20 & 0.700 & \citet{Malaska2025} \\
Chemical energy            & $f_3$ & 0.20 & 0.350 & \citet{Strobel2010} \\
Methane cycle              & $f_4$ & 0.15 & 0.400 & \citet{MitchellLora2016} \\
Surface--atm.\ interaction & $f_5$ & 0.08 & 0.350 & \citet{Miller2021channels} \\
Topographic complexity     & $f_6$ & 0.06 & 0.250 & \citet{Corlies2017Titan} \\
Geomorphological diversity & $f_7$ & 0.04 & 0.300 & \citet{Lopes2019} \\
Subsurface ocean proximity & $f_8$ & 0.02 & 0.030 & \citet{Iess2012} \\
\midrule
\multicolumn{3}{l}{\textit{Global prior mean} $\mu_0$} & 0.331 & \\
\bottomrule
\end{tabular}
\end{table}

\chapter{Geomorphological Organic Abundance Scores}
\label{app:geo_scores}

\begin{table}[H]
\centering
\caption{Organic abundance proxy scores used for the geomorphological
seamless mode of Feature~2.  Grounded in published VIMS spectral
characterisation of each terrain class.}
\label{tab:geo_scores}
\begin{tabular}{llp{5.0cm}c}
\toprule
\textbf{Class} & \textbf{Name} & \textbf{Physical basis} & \textbf{Score} \\
\midrule
Dunes     & Equatorial sand seas & Tholin-dominated; highest VIMS ratio \citep{Lorenz2006}           & 0.82 \\
Plains    & Smooth lowlands      & Organic sediment blanket \citep{Lopes2016}                        & 0.68 \\
Basins    & Topographic lows     & Organic accumulation \citep{Lopes2020}                            & 0.58 \\
Labyrinth & Dissected terrain    & Spectral affinity with plains \citep{Malaska2020}                 & 0.55 \\
Craters   & Impact structures    & Mixed water-ice and organic ejecta \citep{Neish2018}              & 0.35 \\
Mountains & Ridge terrain        & Water-ice-rich crustal material \citep{Solomonidou2018}           & 0.25 \\
Lakes     & Hydrocarbon seas     & Liquid hydrocarbons; low solid organic content \citep{Brown2008}  & 0.05 \\
\bottomrule
\end{tabular}
\end{table}

\chapter{Declared Assumptions and Known Limitations}
\label{app:assumptions}

\begin{enumerate}[leftmargin=*,label=\textbf{A\arabic*.}]

\item \textbf{Conditional feature independence.}  The eight features are
treated as conditionally independent given habitability $H$.  In
practice, organic abundance and chemical energy are positively
correlated, as are liquid hydrocarbon and surface--atmosphere
interaction.  The independence assumption is conservative: correlated
positive features slightly overstate the posterior, but the framework
remains physically motivated and the data-derived importances are
consistent with the prior weights (Section~\ref{sec:importances}).

\item \textbf{Static geomorphological template.}  The Lopes et al.\
(2019) map is used for all five temporal configurations.  The
Late Heavy Bombardment surface had different dune distributions and
crater density, and the future surface will be submerged.  The present
template is used as the best available proxy, with amplitude
modifications applied through epoch-specific prior means and scale
functions.

\item \textbf{Seamless organic abundance mode.}  The organic abundance
feature uses terrain-class scores globally rather than VIMS spectral
band ratios, because the VIMS mosaic covers only $\sim$50\% of
the globe and direct blending produced a visible seam at the coverage
boundary at 180\textdegree{}W.

\item \textbf{Past-epoch Cassini feature contamination.}  The past-epoch
posterior was computed using present-era Cassini SAR maps, which
encode the present polar lake morphology even at $t = -3.5\gya$.
The north polar region is consequently inflated to
$P(H)\approx0.70$ in the past anchor, approximately double the
physically expected value.  This is mitigated in the multi-epoch
interpolation and documented explicitly in the Results.

\item \textbf{Multi-epoch interpolation accuracy.}  PCHIP interpolation
between anchor posteriors is not equivalent to independent
epoch-specific inference.  Intermediate maps correctly reproduce
the anchors and are monotonic between them, but spatial patterns
reflect blended anchor structures.  Estimated uncertainty:
$\Delta P(H) \lesssim \pm0.05$.

\item \textbf{SAR coverage gap.}  Approximately 33\% of \titan{}'s
surface lacks SAR coverage.  Features 1, 3, and 8 are degraded in
these regions, where geomorphological and topographic proxies
substitute.

\end{enumerate}

\end{document}
